\documentclass[12pt]{article}
\special{papersize=210mm,297mm}
\usepackage[top=1.5cm,bottom=1.5cm,left=2cm,right=2cm]{geometry}
\usepackage[skip=12pt plus1pt, indent=0pt]{parskip}

% Usual packages
\usepackage{graphicx}
\usepackage{hyperref}

\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{eurosym}
\usepackage{multicol}
\usepackage{rotate}
\usepackage{tabularx}
\usepackage{color}
\usepackage{colortbl}
\setlength{\headheight}{15.2pt}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0.5pt}
\renewcommand{\footrulewidth}{0.5pt}

\fancyhf{}
%%%% colors
\definecolor{orange}{RGB}{250,167,12}
\definecolor{yellow}{RGB}{246,250,12}
\definecolor{green}{RGB}{128,238,1}
\definecolor{green2}{RGB}{0,250,154}
\definecolor{black}{RGB}{0,0,0}
\definecolor{blue}{RGB}{0,0,255}
\definecolor{red}{RGB}{255,0,0}
\definecolor{white}{RGB}{255,255,255}
\definecolor{burlywood1}{RGB}{255,211,155}
\definecolor{chocolate1}{RGB}{255,127,36}
\definecolor{sepia}{RGB}{94,38,18}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\sepia}[1]{\textcolor{sepia}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\newcommand{\yellow}[1]{\textcolor{yellow}{#1}}
\newcommand{\orange}[1]{\textcolor{orange}{#1}}
%%%% mathematical definitions
\DeclareMathOperator{\Tr}{Tr}
\def\npab{\noindent \textbullet ~}
\def\npa{\noindent}
\def\npat{\noindent \textcolor{blue}{$\blacktriangleright$} ~}
\def\npac{\noindent $\circledast$ ~}
\def\nn{{\bf \nabla}}
\def\df{{\rm d}}
\def\cro{\times}
\def\ip{i^{'}}
\def\jp{j^{'}}
\def\kp{k^{'}}
\def\deg{$^{\circ}$}
\def\rhoc{\textcolor{red}{\rho}}
\def\uic{\textcolor{red}{u_i}}
\def\ujc{\textcolor{red}{u_j}}
\def\pgc{\textcolor{red}{P_{\rm g}}}
\newcommand{\ve}[1]{{\rm\bf {#1}}}
\newcommand{\Fconv}{F_{\rm conv}}
\def\ez{\ve{e}_z}
\def\ey{\ve{e}_y}
\def\ex{\ve{e}_x}
\def\ezero{\ve{e}_{0}}
\def\eplu{\ve{e}_{+1}}
\def\emin{\ve{e}_{-1}}
\def\definition{:=^{\!\!\!\!\!\!\!\textrm{def}}}
\newcommand{\bg}[1]{{\boldsymbol {#1}}}
\newcommand*\tavg[1]{%
  \hbox{%
    \vbox{%
      \hrule height 0.5pt % The actual bar
      \kern0.5ex%         % Distance between bar and symbol
      \hbox{%
        \kern-0.1em%      % Shortening on the left side
        \ensuremath{#1}%
        \kern-0.1em%      % Shortening on the right side
      }%
    }%
  }%
}
\title{Theoretical Astrophysics: Physics of Sun and Stars\\
Homework 4}
\author{P. K\"{a}pyl\"{a}, I. Mili\'{c}}
\date{\today}
%%%%
\begin{document}
\maketitle

\textbf{Deadline for this homework is \textbf{28}/07 23:59}

{\bf Problem 1:} Following the numerical exercise that we did on the
handson number 9, related to stellar atmospheres, calculate the
emergent flux from a given atmospheric model. You can assume that the
optical depth given in the atmospheric model is representative of the
average optical depth over all wavelengths (so-called gray
atmosphere). This means that you do not have to worry about wavelengt
dependence of opacity and thus optical depth.

For the same atmosphere, check whether the flux is constant with
height/depth, and what is the ratio of the K-integral and the mean
intensity with depth.

For this you will need to solve radiative transfer equation
numerically in different directions (you specify a grid of directions)
and to integrate, numerically, intensity over the angle.

{\bf Problem 2:} For a gas of given pressure and temperature, assuming
it only consists of hydrogen in neutral and ionized state, find the
electron density (you will need Saha equation for this, follow
exercise 8).

Assuming that the number density of negative ion of Hydrogen is
much smaller than the number densities of neutral hydrogen and protons, use Saha
equation to estimate the number density of negative ions of hydrogen. By ``much smaller'' we mean that consideration of negative hydrogen ion, does not ``spoil'' the solution of the previous part. 

Now, using Boltzmann equation, estimate the number density of neutral
Hydrogen in state $i=3$ of neutral Hydrogen. Recall that the energy of
the hydrogen states scales like:
\begin{equation}
E_i = \frac{-E_{ion}}{i^2}
\end{equation}
where $i$ is the principal quantum number, $E_i$ is the energy of the
level (negative, because we need to add the energy to break up the
bound state), and $E_{\rm ion}$ is the ionization energy of Hydrogen,
which is equal to 13.6\,eV. Ioniozation energy of the negative ion of
Hydrogen is 0.75\,eV.

Finally, compare the number densities of neutral hydrogen in state
$i=3$ and the negative ion of hydrogen in the model atmosphere you
used in the problem 1. For this problem you will need temperature
(third column) and total gas pressure (fourth column). Keep in mind
the model is given in CGS units! :-)

{\bf Problem 3:} Compute the speed of sound using the formula derived
on lecture 11 in the solar photosphere and at the base of the solar
convection zone. Hint: make use of tabulated values on lecture
4. Compute the Mach number
\begin{equation}
  {\rm Ma} = \frac{u}{c_{\rm s}},
\end{equation}
where $u$ is a typical flow velocity for the same cases. Use $u =
2$~km~s$^{-1}$ in the photosphere and $u=10$~m~s$^{-1}$ at the base of
the convection zone and where $c_{\rm c}$ is the sound speed. Assume
that the gas obeys the ideal gas law.

Do the same analysis for the Earth's atmosphere at sea level. What is
the Mach number in a hurricane with winds $300$~km~h$^{-1}$?


{\bf Useful physical constants}
\begin{itemize}
  \item $R_{\odot} = 696 \times 10^6\,{\rm m}$
  \item $M_{\odot} = 1.989 \times 10^{30}\,{\rm kg}$
  \item $L_{\odot} = 3.83 \times 10^{26}$~W
  \item $T^{\rm eff}_{\odot} = 5777\,{\rm K}$
  \item $1\,{\rm AU} = 1.496 \times 10^8\,{\rm km}$
  \item $c = 2.997 \times 10^8\,{\rm m/s}$
  \item $G = 6.674 \times 10^{-11}$~Nm$^2$/kg$^2$
  \item $k = 1.38\cdot10^{-23}$~J/K
  \item $m_{\rm H} = 1.67\cdot10^{-27}$~kg
  \item $h=6.626 \times 10^{-34}$~J~s.
  \item $k=1.38 \times 10^{-23}$~J/K.
\end{itemize}
\end{document}
