\documentclass[12pt]{article}
\special{papersize=210mm,297mm}
\usepackage[top=1.5cm,bottom=1.5cm,left=2cm,right=2cm]{geometry}
\usepackage[skip=12pt plus1pt, indent=0pt]{parskip}

% Usual packages
\usepackage{graphicx}
\usepackage{hyperref}

\usepackage{fancyhdr}
\usepackage{amsmath}
\usepackage{cancel}
\usepackage{amssymb}
\usepackage{eurosym}
\usepackage{multicol}
\usepackage{rotate}
\usepackage{tabularx}
\usepackage{floatrow}
\usepackage{color}
\usepackage{colortbl}
\usepackage{braket}
\usepackage[shortcuts]{extdash}
\setlength{\headheight}{15.2pt}
\pagestyle{fancy}
\renewcommand{\headrulewidth}{0.5pt}
\renewcommand{\footrulewidth}{0.5pt}

\fancyhf{}
%%%% colors
\definecolor{orange}{RGB}{250,167,12}
\definecolor{yellow}{RGB}{246,250,12}
\definecolor{green}{RGB}{128,238,1}
\definecolor{green2}{RGB}{0,250,154}
\definecolor{black}{RGB}{0,0,0}
\definecolor{blue}{RGB}{0,0,255}
\definecolor{red}{RGB}{255,0,0}
\definecolor{white}{RGB}{255,255,255}
\definecolor{burlywood1}{RGB}{255,211,155}
\definecolor{chocolate1}{RGB}{255,127,36}
\definecolor{sepia}{RGB}{94,38,18}
\newcommand{\blue}[1]{\textcolor{blue}{#1}}
\newcommand{\sepia}[1]{\textcolor{sepia}{#1}}
\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\green}[1]{\textcolor{green}{#1}}
\newcommand{\yellow}[1]{\textcolor{yellow}{#1}}
\newcommand{\orange}[1]{\textcolor{orange}{#1}}
%%%% mathematical definitions
\DeclareMathOperator{\Tr}{Tr}
\def\npab{\noindent \textbullet ~}
\def\npa{\noindent}
\def\npat{\noindent \textcolor{blue}{$\blacktriangleright$} ~}
\def\npac{\noindent $\circledast$ ~}
\def\nn{{\bf \nabla}}
\def\df{{\rm d}}
\def\cro{\times}
\def\ip{i^{'}}
\def\jp{j^{'}}
\def\kp{k^{'}}
\def\deg{$^{\circ}$}
\def\rhoc{\textcolor{red}{\rho}}
\def\uic{\textcolor{red}{u_i}}
\def\ujc{\textcolor{red}{u_j}}
\def\pgc{\textcolor{red}{P_{\rm g}}}
\newcommand{\ve}[1]{{\rm\bf {#1}}}
\newcommand{\Fconv}{F_{\rm conv}}
\def\ez{\ve{e}_z}
\def\ey{\ve{e}_y}
\def\ex{\ve{e}_x}
\def\ezero{\ve{e}_{0}}
\def\eplu{\ve{e}_{+1}}
\def\emin{\ve{e}_{-1}}
\def\definition{:=^{\!\!\!\!\!\!\!\textrm{def}}}
\newcommand{\bg}[1]{{\boldsymbol {#1}}}
\newcommand*\tavg[1]{%
  \hbox{%
    \vbox{%
      \hrule height 0.5pt % The actual bar
      \kern0.5ex%         % Distance between bar and symbol
      \hbox{%
        \kern-0.1em%      % Shortening on the left side
        \ensuremath{#1}%
        \kern-0.1em%      % Shortening on the right side
      }%
    }%
  }%
}
\title{Theoretical Astrophysics: Physics of Sun and Stars\\
Homework 2}
\author{P. K\"{a}pyl\"{a}, I. Mili\'{c}}
\date{\today}
%%%%
\begin{document}
\maketitle

\textbf{Deadline for this homework is 04/06 23:59}

{\bf Problem 1:} The solar luminosity is $L_\odot = 3.83\cdot
10^{26}$~W. Assume that all of the energy for this luminosity is
provided by the $pp1$ chain, and that neutrinos carry off 3\% of the
energy. How many neutrinos are produced per second? What is the
neutrino flux (i.e.\ the number of neutrinos per second per cm$^2$) at
Earth?

The $pp1$ chain is comprised of the reactions:
\begin{eqnarray}
  p + p &\rightarrow& ^2{\rm D} + e^+ + \nu,\ (1.18~{\rm MeV})\nonumber \\
  ^2{\rm D} + p &\rightarrow& ^3{\rm He} + \gamma,\ \ \ \ \ \ (5.49~{\rm MeV}) \nonumber \\
  ^3{\rm He} + ^3{\rm He} &\rightarrow& ^4{\rm He} + 2p.\ \ \ \ \ (12.86~{\rm MeV})
\end{eqnarray}
where the first two reactions need to happen twice for the last
reaction to occur, and where $\nu$ signifies an emitted
neutrino. Energy released (in MeV = $10^6$~eV) is given in brackets
after each reaction.


{\bf Problem 2:} a) Why does convection transport heat radially outward
although there is no net mass flux?

b) Estimate the superadiabatic temperature gradient in order of
magnitude fashion in the Sun making use of the mixing length
expression for $\Fconv$:
\begin{equation}
  \Fconv = c_p \rho T g^{1/2} \frac{\ell^2}{4\sqrt{2}H_p^{3/2}} (\nabla-\nabla_{\rm ad})^{3/2}.
\end{equation}
Use the average values of density and temperature of the Sun, and $c_p
= 2.07\cdot10^4$~J~K$^{-1}$~kg$^{-1}$. Hint: Recall that $\ell =
\alpha_{\rm MLT}H_p$ and bear in mind the definition of the pressure
scale height from previous homework. When is this result a good
approximation and when not? Why?


{\bf Problem 3:} Use the equations of hydrostatic equilibrium and mass
conservation:
\begin{eqnarray}
  \frac{dp}{dr} = -\rho \frac{Gm}{r^2},\ \ \frac{dm}{dr} = 4\pi r^2\rho, 
\end{eqnarray}
assuming a polytropic equation of state
\begin{equation}
p = K \rho^\gamma,
\end{equation}
where $\gamma = 1 + \frac{1}{n}$ is the polytropic exponent and $n$ is
the polytropic index, to derive the Lane-Emden equation:
\begin{equation}
\frac{1}{\xi^2} \frac{d}{d\xi} \left(\xi^2 \frac{d\theta}{d\xi} \right) = -\theta^n.
\end{equation}
Here $\theta$ and $\xi$ are the non-dimensional density and radius
defined as
\begin{equation}
\rho = \rho_{\rm c} \theta^n, \ \ \mbox{and}\ \ r = \alpha\xi,
\end{equation}
and where $\alpha^2$ equals a constant that arises in the derivation
of the equation.

The boundary conditions at $\xi=0$ for the Lane-Emden equation are:
\begin{equation}
\theta = 1,\ \mbox{and}\ \frac{d\theta}{d\xi} = 0.
\end{equation}
Solve the Lane-Emden equation for a few values of $n$ in the range
$n=[0,10]$. You will need to do this numerically because analytical
solutions exists only for $n=0$, $1$, and $5$.


{\bf Useful physical constants}
\begin{itemize}
  \item $R_{\odot} = 696 \times 10^6\,{\rm m}$
  \item $M_{\odot} = 1.989 \times 10^{30}\,{\rm kg}$
  \item $L_{\odot} = 3.83 \times 10^{26}$~W
  \item $T^{\rm eff}_{\odot} = 5777\,{\rm K}$
  \item $1\,{\rm AU} = 1.496 \times 10^8\,{\rm km}$
  \item $c = 2.997 \times 10^8\,{\rm m/s}$
  \item $G = 6.674 \times 10^{-11}$~Nm$^2$/kg$^2$
  \item $k = 1.38\cdot10^{-23}$~J/K
  \item $m_{\rm H} = 1.67\cdot10^{-27}$~kg
  \item $h=6.626 \times 10^{-34}$~J~s.
  \item $k=1.38 \times 10^{-23}$~J/K.
\end{itemize}
\end{document}
